Visually pleasing knot projections

### 2021-09-26 inst/jma_knot.tex
fixes #39

It is common to present a knot using diagrams such as
Figure~\ref{knot_table}, in which a two-dimensional projection of the
knot is given with broken lines indicating where one strand passes
over another.  Knot diagrams have a long history:
Przytycki~\cite{przytycki1998} discusses an intricate braid used as a
seal sign in around 1700\,BCE; they figure prominently in {\em The
  Book of Kells} and the Lindisfarne gospels, graphic art produced
around 800\,CE~\cite{bain1973}.  Such artwork makes extensive use of
symmetry~\cite{cromwell2008}.

Knot diagrams are ubiquitous in the modern discipline of mathematical
knot theory.  In the nineteeth century we see Tait~\cite{tait1884}
publishing an early systematic study of knots including a beautiful
plate of elegantly hand-drawn cursive knot diagrams.  The tradition
continues into the twentieth century with Rolfsen~\cite{rolfsen1976}
presenting an appendix containing many knot and link projections.
Rolfsen's diagrams are used widely today: Adams~\cite{adams2004}
reproduces his work extensively, as do many more recent works.  Online
resources are becoming more widely available, with one prominent
example being the {\em Knot Atlas}~\cite{knot_atlas} presenting
extensive knot tables following Rolfsen~\cite{rolfsen1976}---for knots
up to~10 crossings---and Thistlethwaite~\cite{hoste1998}, for knots
of~11 crossings.  Symbolic software such as Mathematica and Maple can
be used to produce representations of knots, typically in the form of
nonstandard embeddings of the torus into~$\mathcal{R}^3$; but as
discussed below this approach is suboptimal from the perspective of producing a
visually appealing projection.

@@

The first step is to create a closed curve in inkscape that shows the
rough outline of the knot~(Figure~\ref{screenshot_inkscape_7_6} shows
a screenshot of {\tt 7\_6\_first\_draft.svg}, supplied with the
package).  Note that this file contains only the knot {\em path}; the
over and under information is to be added later.

@@

Many of the knots in Figure~\ref{knot_table} have an axis of symmetry,
or possess rotational symmetry; some have both and thus have a
dihedral symmetry group.  The package has the ability to impose these
symmetry relations on knots, and to optimize the resulting symmetrical
knot.  Minimizing the badness is not entirely straightforward on
account of the induced redundancy, which is characterized using a
symmetry object specific to the knot under consideration.  However,
symmetry constraints do reduce the dimensionality of the optimization
problem.

>>>>

It is common to present a knot using diagrams such as
Figure~\ref{knot_table}, in which a two-dimensional projection of the
knot is given with broken lines indicating where one strand passes
over another.  Knot diagrams have a long history:
Przytycki~\cite{przytycki1998} discusses an intricate braid used as a
seal sign in around 1700\,BCE; they figure prominently in {\em The
  Book of Kells} and the Lindisfarne gospels, graphic art produced
around 800\,CE~\cite{bain1973}.  Such Celtic artwork makes extensive
use of symmetry~\cite{cromwell2008}, and their visual appeal persists
to the present day~\cite{antonsen2021}.  More recent examples would
include bobbin lace~\cite{irvine2020}, in which many threads are
braided together to form a loose fabric.  The enduring popularity of
such fibre arts attests to the aesthetic of knots and braids in the
modern world

Knot diagrams are ubiquitous in the modern discipline of mathematical
knot theory.  In the nineteeth century we see Tait~\cite{tait1884}
publishing an early systematic study of knots including a beautiful
plate of elegantly hand-drawn cursive knot diagrams.  The tradition
continues into the twentieth century with Rolfsen~\cite{rolfsen1976}
presenting an appendix containing many knot and link projections.
Rolfsen's diagrams are used widely today: Adams~\cite{adams2004}
reproduces his work extensively, as do many more recent works.  Online
resources are becoming more widely available, with one prominent
example being the {\em Knot Atlas}~\cite{knot_atlas} presenting
extensive knot tables following Rolfsen~\cite{rolfsen1976}---for knots
up to~10 crossings---and Thistlethwaite~\cite{hoste1998}, for knots
of~11 crossings.

\subsection{Computer-generated knot visualisations}

Symbolic software such as Mathematica and Maple can be used to produce
representations of knots, typically in the form of nonstandard
embeddings of the torus into~$\mathbb{R}^3$; but as discussed below
this approach is suboptimal from the perspective of producing a
visually appealing projection.

Visual appeal of knot projections appears to be important to
contemporary mathematicians.  Recent work by Taalman (writing as
mathgrrl)~\cite{taalman2019}, for example, shows that large amounts of
effort have been expended by serious mathematicians making
representations of knots attractive.  Robert Bosch~\cite{bosch2010},
taking a different approach to knot representations, presents artwork
that utilises high-dimensional computer optimization in a similar way
to that presented here.

No discussion of computer-generated representations of knots would be
complete without the tour de force that is~{\em
  KnotPlot}~\cite{scharein1997,scharein1998} which includes extensive
functionality for visualising knots.  Much of the visualization
considers (nonstandard) embeddings of a torus in $\mathbb{R}^3$; the
configuration is ``relaxed'' in such a way as to produce a smoother
output.  {\em KnotPlot} has two major display modes: any knot may be
displayed in either a ``beads and sticks" mode or in a ``smooth tubes"
mode.  The software does render knot diagrams but the author states
[p77] that ``knots with higher crossing numbers (greater than six or
seven) will not produce good 2D projections when fully relaxed"; he
goes on to say that this phenomenon merits further study, and admits
that some of the diagrams are ``a bit messy".

%One of the early works using this approach is the 1932 German classic
%Anschauliche Geometrie by David Hilbert and S. Cohn-Vossen, available
%in English translation as Geometry and the Imagination [HCV52].  This
%book is remarkable for the quality of insight and also for its
%incredible diagrams, mostly hand drawn and (obviously) predating the
%computer era.

\subsubsection{Knot theory in R}

The R programming language~\cite{rcore2021}, while usually associated
with statistical analysis and data processing, is also a suitable tool
for producing visually attractive illustrations.  Generative art [that
  is, use of an algorithm, with or without a random component, in a
  creative process] is possible in R~\cite{brunner2021}.  More
specficially to knot theory, the {\tt Rknots}
package\cite{comoglio2011,comoglio2016} presents software for folded
protein structures, and includes 3D renderings of commonly encountered
knotted strands.

\section{The knotR package: Rationale}

@@

The first step is to create a closed curve in inkscape that shows the
rough outline of the knot~(Figure~\ref{screenshot_inkscape_7_6} shows
a screenshot of {\tt 7\_6\_first\_draft.svg}, supplied with the
package).  Note that this file contains only the knot {\em path}; the
over and under information is to be added later.  Irvine {\em
  et. al.}~\cite{irvine2020} use the terms {\em drawing} and {\em
  braid word} for analogous concepts in their work on bobbin lace
patterns.

@@

If a knot diagram can be drawn with a particular symmetry, one common
response is to demand that this symmetry be respected exactly.
Bosch~\cite{bosch2010}, for example, considered rotational symmetry to
be sufficiently important to impose five-fold rotatioal symmetry in
one of his works; similar comments apply to mirror symmetry.  Many of
the knots in Figure~\ref{knot_table} have an axis of symmetry, or
possess rotational symmetry; some have both and thus have a dihedral
symmetry group.  The package implements Bosch-type symmetry: one can
impose these symmetry relations on knots, and optimize the resulting
symmetrical knot.

Minimizing the badness is not entirely straightforward on account of
the induced redundancy, which is characterized using a symmetry object
specific to the knot under consideration.  However, symmetry
constraints do reduce the dimensionality of the optimization problem.
